---
title: 'Review: Sampling Distributions and Confidence Intervals'
header-includes: \usepackage{amssymb}
output:
  beamer_presentation:
    colortheme: seahorse
    fonttheme: default
    keep_tex: yes
    template: ../../beamer169.tex
  ioslides_presentation: default
fontsize: 12pt
---

```{r global_options, include=FALSE, echo=FALSE}
.pardefault <- par()
# use for beamer
knitr::opts_chunk$set(fig.width=4, fig.height=3, fig.align='center',warning=FALSE, message=FALSE)
# use for word
# knitr::opts_chunk$set(fig.width=4, fig.height=3,warning=FALSE, message=FALSE)
source('../scripts/shadeMe.R')

# install DS705data if needed
if (!is.element("DS705data",installed.packages()[,1]))
{ install.packages("http://websites.uwlax.edu/jbaggett/DS705/DS705data_1.0.tgz",repos=NULL,type="source")}

```

## Bootstrap Methods

\Large

Useful when

- small sample
- or conditions not met
- or no procedure available

Idea: simulate the sampling distribution

<div class="notes">
- these notes are not intended to be a comprehensive introduction to bootstrapping, rather a brief introduction to illustrate how simulating a sampling distribution can be useful
</div>

## Bootstrap Methods for $\mu$

\Large

- random variable is not normally distributed
- $n$ is small

<div class="notes">
- The $t$-test can handle some departure from normality especially for very large samples.  
- But as we saw with the test-statistic distribution shiny app.  Small samples and strong non-normality result in a $t$ test-statistic whose distribution departs from a $t$ distribution.
- The bootstrap is one, of several, alternative methods that can be used when the exact $t$ procedures are inappropriate
</div>


## Bootstrap Idea

\Large

- original sample is the "pseudo" population

- resample with replacement

- $\Rightarrow$ approximate sampling distribution


## Men's Cholesterol Levels

\large $n = 40$ and *skewed* distribution

```{r, echo=FALSE, message=FALSE, fig.width=4, fig.height=2.5}
require(DS705data);
data(HealthExam)
c <- HealthExam$Cholesterol[HealthExam$Sex=="M"]
xbar <- round(mean(c),digits=2)
s <- round(sd(c),digits=2)
par(.pardefault)
llim <- 0
ulim <- 1300
par(mar=(c(4,.5,0,1)),yaxs="i",bty="l",xaxs="i",yaxt="n")
hist(c,main=" ",xlab="mg",prob=TRUE,ylab="")
abline(h=0,col="black",lwd=2)
n = length(c)
```

## Bootstrap CI simulated $t$ distribution

```{r,echo=FALSE, fig.width=5,fig.height=3}
chol <- HealthExam$Cholesterol[HealthExam$Sex=="M"]
n <- length(chol)
xbar <- mean(chol)
s <- sd(chol)

reps <- 10000
t_sim <- numeric(reps)

# generate bootstrapped sampling distribution
for (i in 1:reps){
  x <- sample(chol,n,replace=TRUE)
  t_sim[i] <- (mean(x)-xbar)/(sd(x)/sqrt(n))
}

# build boostrapped "Studentized" t-inverval for mu
C <- .95
alpha <- 1 - C
prbs <- c( 1-alpha/2, alpha/2 )

tcrit_est <- unname(quantile(t_sim, prob = prbs ))
t_ci_bootstrap <- xbar - tcrit_est*s/sqrt(n)

# for comparison here is the t-interval from theory
t_ci_theory <- as.vector(t.test(chol)$conf.int)


par(mar=(c(4,.5,0,.5)),yaxs="i",bty="l",xaxs="i",yaxt="n")
brks <- seq(min(t_sim),max(t_sim),length=50)
hist(t_sim,prob=TRUE,breaks=brks,main='',xlab='t',ylab='')
x0 <- seq(-4,4,length=400)
y0 <- dt(x0,df=39)
points(x0,y0,type='l',lwd=2,col='blue')
```

<div class="notes">
- Notice that the sampling distribution of t is skewed left because of the skew in the original data.
- The theoretical critical values would be inaccurate.
</div>

## The critical values

\Large

critical value | bootstrap | theory
--- | --- | ---
$t_{.025}$ | `r signif(tcrit_est[2],3)` | `r signif(qt(.025,df=39),3)`
$t_{.975}$ | `r signif(tcrit_est[1],3)` | `r signif(qt(.975,df=39),3)`

<div class="notes">
- Notice that the bootstrap critical values are asymmetrical while the theoretical critical values are symmetric
- the bootstrap critical values correctly reflect the skewness of the original data
</div>

## Unfortunate Notation

\Large
$t_{.025}$

- the .025 or the .975 percentile?
- context
- if symmetry, then not important

## Formula for the Confidence Interval

\Large

$$ t_{.025} < t = \frac{\overline{x} - \mu}{s / \sqrt{n}} < t_{.975} $$

$$ \overline{x} - t_{.975} \frac{s}{\sqrt{n}} < \mu < \overline{x} - t_{.025} \frac{s}{\sqrt{n}} $$

Formula in Ott is wrong, compare above to page 260

## Results

\Large
```{r echo=FALSE}
xbar <- round(xbar,2)
mbs1 <- round(xbar-t_ci_bootstrap[1],2)
mbs2 <- round(t_ci_bootstrap[2]-xbar,2)
mt1 <- round(xbar-t_ci_theory[1],2)
mt2 <- round(t_ci_theory[2]-xbar,2)
```

Bootstrapped (good):
$$ (`r xbar` - `r mbs1`, `r xbar` + `r mbs2`) = (`r xbar-mbs1`,`r xbar+mbs2`) $$

Theory (do not use): 
$$ (`r xbar` - `r mt1`, `r xbar` + `r mt1`) = (`r xbar-mt1`,`r xbar+mt1`) $$

<div class="notes">
- note that the bootstrapped interval is not centered about the sample mean but is instead longer to the right and shorter to the left ...
- this reflects the same skewness as the orginal data
- Since the original data seems to be quite skewed and the sample size is small it would be better to report the results of the bootstrapped ci
- So.  We are 95% confident the population mean men's cholesterol level is between XXX and XXX
</div>

<!--
## Bootstrap Hypothesis Test for $\mu$

\large
Is average cholesterol level for U.S. men larger than 320 mg?

$$ H_0: \mu = 320, \hspace H_1: \mu > 320 $$

$t$-test not accurate
-->

## Results of the $t$-test for $\mu$

```{r}
t.test( chol, mu = 320, alternative="greater")
```


## Bootstrap $t$-test

Simulate test statistic with 10,000 resamples:

```{r,echo=FALSE, fig.width=5,fig.height=3}
n <- length(chol)
xbar <- mean(chol)
s <- sd(chol)
mu0 <- 320
t0 <- (xbar-mu0)/(s/sqrt(n))
p0 = pt(t0,df = n-1, lower.tail=FALSE)

reps <- 10000
t_sim <- numeric(reps)


for (i in 1:reps){
  x <- sample(chol,n,replace=TRUE)
  t_sim[i] <- (mean(x)-xbar)/(sd(x)/sqrt(n))
}

p1 <- signif(sum(t_sim>t0)/reps,3)

x <- seq(min(t_sim),max(t_sim),length=400)
y <- dt(x,df=n-1)
brks <- seq(min(t_sim),max(t_sim),length=50)
par(mar=(c(4,.5,0,.5)),yaxs="i",bty="l",xaxs="i",yaxt="n")
hist( t_sim, prob=TRUE, breaks=brks,main='',xlab='t',ylab='')
points(x,y,type = 'l',col='blue',lwd=2)
points(c(t0,t0),c(0,dt(t0,df=39)),type='l',col='red',lwd=2)
text(2.2,.2,bquote(P%~~%.(p1)))        
```

## Result

\large

- Cholesterol levels very skewed $\Rightarrow$ $t$-test inaccurate.

- Bootstrap $t$-test accounts for the skewness so reject $H_0$

- *Men's average cholesterol level is greater than 320 mg ($P$ = `r p1`)*