---
title: "Using the boot package"
output:
  beamer_presentation:
    colortheme: default
    fonttheme: default
    keep_tex: yes
    template: ../../beamer169experimental.tex
fontsize: '12pt'
---

## Powerful, but not easy

The "boot" package is powerful, but has a learning curve!

```{r}
require(boot)  # install 'boot' if needed
```

## Statistic function
 
To use `boot()` you have have to write an auxiliary function of two inputs that computes the statistic you want to bootstrap.

```{r}
bootMedian <- function( x, i){
  # x is a vector of data
  # i is a vector of indices for the resampled
  #   elements of x
  median(x[i]) # gets returned
}
```



## Statistic function (2)

Here is a version with a print statement so we can see how it works.
```{r}
bootMedianPrint <- function( x, i){
  # x is a vector of data
  # i is a vector of indices for the resampled
  #   elements of x
  print(i)
  print(x[i])
  median(x[i]) # gets returned
}
```

## How it works

```{r}
set.seed(1)
x <- c(.2,.3,.5,.7,1.1)
boot.tmp <- boot(x,bootMedianPrint,R=2)
```

`boot.tmp` stores the resampled statistic values and other stuff ...

<div class='notes'>
- the first line shows us the first index vector used by boot, it's 1 2 3 4 5 showing us that boot is first computing the statistic for the original sample
- notice the second line is the 1,2,3,4,5 entries of x
- the third line shows the index vector i = 2 3 2 5 4, remember we are resampling with replacement, so this is telling us to construct a new sample with the 2nd, third, 2nd, fifth, and fourth entries of x 
- the fourth line shows the corresponding entries of x ... the bootMedian function will compute the median of the fourth line and return the result to boot()
- the fifth and sixth lines show the process repeated 
- in an actual bootstrap scenario we'd likely compute thousands of these resamples and bootMedian would return the median of each resample
</div>

## Bootstrapping the median

```{r,fig.height=2,fig.width=4,fig.align='center'}
x <- c(2,7,7,8,11,13,15,33) # the original sample
set.seed(NULL); boot.object <- boot(x,bootMedian,R=5000)
par(mar=c(4.5,0,0,0))
hist(boot.object$t,breaks=40,main='t',xlab='sample median',yaxt='n')
```
<div class='notes'>
the bootstrap statistic is stored in the boot.object returned by boot().
To see what all is stored in the list boot.object you can apply the str() or structure command to the boot.object.
</div>

## Some things to try

You try:
```{r eval = FALSE}
# see everything saved in the results of boot
str(boot.object)
# plot a boot object
plot(boot.object)
```

## Bootstrap confidence intervals

Build confidence intervals from the bootstrap object

```{r, warning=F,eval=FALSE}
boot.ci(boot.object)
````

```{r, echo=FALSE,warning=FALSE}
require(utils)
tmp <- noquote(capture.output(boot.ci(boot.object)))
write.table(tmp[1:5],quote=F,row.names=F,col.names=F)
```


## Bootstrap confidence intervals (continued output)
```{r,echo=FALSE}
write.table(tmp[6:13],quote=F,row.names=F,col.names=F)
```

<div class='notes'>
- there are multiple ways to compute a confidence interval from a bootstrap distribution
- the one to focus on in this list is the 'bca' interval ... this is the Bias-Corrected Accelerated bootstrap and it adjusts for bias and skewness in the bootstrap distribution.  Of the intervals listed here, it is usually the most accurate.
- one interval that isn't listed here is the Studentized interval like those you studied in Week 2 of this class
- to get the boot.ci command to produce Studentized intervals the auxiliary function has to return both the statistic and the estimated standard error of the statistic which isn't always easy to do.  
- So even though Studentized intervals are optimal in some ways, the bca intervals are good and easier to compute from a users perspective.
- Sometimes you'll get error messages when trying to compute bca intervals because of the underlying complex calculations.  If the results are suspect or very different than the other results, you can report one of the other intervals, but read up on the other intervals so you know what you're reporting.  Don't use the normal ci, unless you've got large samples or data from a distribution close to normal.
</div>

## Bootstrap Info

\Large 

- Requirements
    - \large sample must be representative of population
    - very small samples are still a problem
- Advantage
    - \large can bootstrap estimate any statistic
    
## Bootstrap trimmed mean

```{r}
bootTrMean <- function( x, i, trim=0){
  mean(x[i],tr=trim) # gets returned
}
set.seed(43);
x <- rlnorm(50); # very skewed distribution
boot.object <- boot(x,bootTrMean,R=5000,trim=0.2)
```

## Bootstrap trimmed mean

```{r}
boot.ci(boot.object,type='bca')
```

## Bootstrap variance

Estimate population variance (hint: it's 25)

```{r}
bootVar <- function( x, i){
  var(x[i]) # gets returned
}
x <- rnorm(40,mean=10,sd=5)
boot.object <- boot(x,bootVar,R=5000)
```

## Bootstrap variance (2)

```{r}
boot.ci(boot.object,type='bca')
```

## Bootstrap difference using strata

First we need to set up our auxiliary stat function

```{r}
bootMeanDiff <- function(d,i){
  # d is a dataframe, d[,1] is data, d[,2] is group factor variable
  # d[i,1] is resampled data
  print(d[i,1]) # remove or comment out for general use
  means <- tapply(d[i,1],d[,2],mean)
  means[1]-means[2]
}
```

## Bootstrap difference using strata (2)

```{r}
x <- c(1.1, 1.3, 1.4, 1.8, 2.2, 2.5, 2.6, 2.7)
group <- factor(rep(c('A','B'),c(4,4)))
d <- data.frame(x,group)
boot.object <- boot(d, bootMeanDiff, R = 5, strata = d$group )
```

<div class='notes'>
- specifying the strata option by setting strata equal to your factor variable
ensure that the resampling occurs only within the groups
- to see what I mean, look at the output, each resample consists of 4 elements of the A group followed by 4 elements of the B group
- that way the auxiliary function can compute the difference of means between the groups
</div>

## Difference of means example

```{r echo=FALSE}
bootMeanDiff <- function(d,i){
  means <- tapply(d[i,1],d[,2],mean)
  means[1]-means[2]
  }
```

```{r}
x <- c(rnorm(30,0,1),rnorm(30,5,4))
group <- factor(rep(c('A','B'),c(30,30)))
d <- data.frame(x,group)
boot.object <- boot(d, bootMeanDiff, R = 5000, strata = d$group)
```

<div class='notes'>
we removed the print statement from the  function before running the bootstrap
</div>

## Difference of means example (2)

```{r}
boot.ci(boot.object,type='bca')$bca[4:5]
xA <- d$x[d$group=='A']; xB <- d$x[d$group=='B'] 
t.test(xA,xB)$conf.int
```

<div class='notes'>
- wow, the bootstrap approximated ci is very close to the parametric ci produced by a t-procedure.  
- Since we sampled from  normal distributions this is expected.  
- note, the actual difference between the population means is -5 which is nicely captured in both intervals
</div>

## Multiple Differences


```{r, echo=TRUE, eval = FALSE}
x <- c(rnorm(30,0,1),rnorm(20,2,2),rnorm(40,5,4),rnorm(20,6,1))
group <- factor(rep(LETTERS[1:4],c(30,20,40,20)))
d <- data.frame(x,group)
```

```{r, echo=FALSE, fig.height=2, fig.width=4, fig.align='center'}
set.seed(44)
x <- c(rnorm(30,0,1),rnorm(20,2,2),rnorm(40,5,4),rnorm(20,6,1))
group <- factor(rep(LETTERS[1:4],c(30,20,40,20)))
d <- data.frame(x,group)
par(mar=c(2,0,0,0)); boxplot(x~group,data=d)
```

## Auxiliary Function

The auxiliary function can returns as many statistics as we like

```{r echo=TRUE}
bootMeanDiff <- function(d,i){
  means <- tapply(d[i,1],d[,2],mean)
  c( means[2]-means[1], means[3]-means[1], means[4]-means[1],
     means[3]-means[2], means[4]-means[2], means[4]-means[3] )
  }
# build the bootstrap distribution
boot.object <- boot(d, bootMeanDiff, R = 5000, strata = d$group)
```

## CI's for differences (1)

```{r echo=TRUE}

# mu_B - mu_A (= 2)
boot.ci(boot.object,conf = 1 - .05/6, type='bca', index=1)$bca[4:5]
# mu_C - mu_A (= 5)
boot.ci(boot.object,conf = 1 - .05/6, type='bca', index=2)$bca[4:5]
# mu_D - mu_A (= 6)
boot.ci(boot.object,conf = 1 - .05/6, type='bca', index=3)$bca[4:5]
```

<div class='notes'>
if the auxiliary function returns a vector of statistics like the one on the previous slide, 
then use the index argument to boot.ci to specify which statistic you want to estimate.
</div>

## CI's for differences (2)

```{r}
# mu_C - mu_B (= 3)
boot.ci(boot.object,conf = 1 - .05/6, type='bca', index=4)$bca[4:5]
# mu_D - mu_B (= 4)
boot.ci(boot.object,conf = 1 - .05/6, type='bca', index=5)$bca[4:5]
# mu_D - mu_C (= 1)
boot.ci(boot.object,conf = 1 - .05/6, type='bca', index=6)$bca[4:5]
```

## CI's for differences (3)
`onewayComp()` also do pairwise Studentized CI's for differences of means
```{r}
source('./onewayComp.R') # or in DS705data package
onewayComp(x~group,data=d,alpha=.05/6,nboot=0,
           var.equal=F, adjust='none')$comp[,2:3]
```

## CI's for differences (4)
`onewayComp()` also has a bootstrap option
```{r}
source('./onewayComp.R')
onewayComp(x~group,data=d,alpha=.05/6,nboot=10000,
           var.equal=F, adjust='none')$comp[,2:3]
```