---
title: 'Day 1 Orientation'
author: "DS705"
fontsize: '12pt'
output:
  beamer_presentation:
    colortheme: "seahorse"
    keep_tex: true
    template: '../../beamer169experimental.tex'
    fonttheme: default
---



```{r global_options, include=FALSE, echo=FALSE}
# use for beamer
knitr::opts_chunk$set(fig.width=4, fig.height=3, warning=FALSE, message=FALSE)
# use for word
# knitr::opts_chunk$set(fig.width=4, fig.height=3,warning=FALSE, message=FALSE)
```

## Course Intro Video

up to Ryan and Bryan to include here or not

## Instructor Intro Video

up to Ryan and Bryan

## R

\large

- *Free*, open source statistical analysis software
- Computational engine for DS705
\begin{center}
\includegraphics[width=4in]{figures/2015-vs-2014-donuts.png}
\end{center}

<div class='notes'>
- put http://www.burtchworks.com/2015/05/21/2015-sas-vs-r-survey-results/ in the box below slide
- a hot topic in the data science community is whether to learn R or SAS, we focus on R here for three reasons
    1. it has packages available to do almost anything statistical
    2. it's poised to overtake SAS as the most common platform
    3. it's free - we have it on the virtual desktop for your use, but you should also install R and Rstudio on your own machine.  I use it as my personal statistical calculator.
</div>

## Recent News about R

\Large

- 2015 - Microsoft acquires commercial R provider Revolution Analytics
- 2016 - In-database R coming to SQL server

<div class='notes'>
- no audio
add links below slide
- Revolution Analytics - http://www.cio.com/article/2906456/data-analytics/microsoft-closes-acquisition-of-r-software-and-services-provider.html
- R in SQL - http://blog.revolutionanalytics.com/2015/05/r-in-sql-server.html
</div>

## RStudio
 
- Development environment for DS705 in virtual desktop
\begin{center}
\includegraphics[width=3in]{figures/RStudio-Screenshot.png}
\end{center}

## RStudio Overview Video

* please embed this video*

- suggest HD and fullscreen
- https://vimeo.com/97166163

## Class Structure (non-exam weeks)

\large

- Study presentation(s)
- Read
- Complete swirl lesson (some weeks)
- Contribute to group discussion
- Complete brief quiz in D2L
- Complete homework using R Markdown and submit via D2L

## Class Structure (exam weeks)

\large

- Work on practice problems
- Take exam
    - \large May include D2L part similar to quizzes
    - Complete exercises in R Markdown and submit via D2L

## swirl

\begin{center}\includegraphics[width=5in]{figures/swirl_logo.png}\end{center}

<div class='notes'>
- Put link to http://swirlstats.com below slide
- We'll use swirl extensively throughout the first half of DS705 to help you learn to about R and to learn how to conduct simple statistical analyses in R
- You're not required to submit any work from swirl, but it will definitely help you with the R part of this class
</div>


## Installing swirl and swirl Course

At the console in the R-studio:
```{r eval=FALSE}
install.packages('swirl') 
library(swirl)
install_course_github('DataScienceUWL','Statistical_Methods') 
swirl() 
info() # type this anytime in swirl to see options
```
After the first time, you can skip the first and third commands.

<div class='notes'>
- no audio
- add code in box below slide for easy cut and paste

install.packages('swirl')  
library(swirl)  
install_course_github('DataScienceUWL','Statistical_Methods')   
swirl()
info() # type this anytime in swirl to see options  

</div>

## R Markdown

\large

- Great tool for reproducible research
- Each week download file like 'Homework_01.Rmd'
- In RStudio
    - \large open .Rmd file
    - Add chunks of R code and text to complete exercises
    - 'knit' a Word .docx file
    - Upload .Rmd and .docx file to D2L dropbox
- http://rmarkdown.rstudio.com for getting started and documentation

<div class='notes'>
- no audio
</div>  

## Try R Markdown Now

\large 

- In RStudio
    - \large Choose File $\rightarrow$ New File $\rightarrow$ R Markdown
    - Fill in the title and author fields and select Word document
    - At the top of the editor window, click on 'Knit Word'
    
## Weekly Download

\large

- Download week01_download.zip from D2L
- Create DS705 directory and unzip there
- Set working directory to week01_download in RStudio
    - Session $\rightarrow$ Set Working Directory $\rightarrow$ Choose Directory
- Open Week_01_Discussion_Assignment.Rmd
- Open Week_01_HW_Submission.Rmd
    