
---
title: 'Lab Assignment: Logistic Regression'
author: "Spencer Swartz"
date: "April 4, 2017"
output: word_document
fontsize: 12pt
---



Create a Word document from this R Markdown file for the following exercises.  Submit the R markdown file and resulting Word document via D2L Dropbox.   

## Exercise 1

A study was conducted whereby the type of anesthetic (A or B), nausea after the surgery (Yes or No), the amount of pain medication taken during the recovery period, and age for a random sample of 72 patients undergoing reconstructive knee surgery.  

The data is in the file anesthesia.rda.

### Part 1a

Use R to create a two-way table with the type of anesthetic defining the rows and nausea after the surgery as the columns and also produce the output for a chi-square test for independence.

Is there an association between these two categorical variables at a 5% level of significance?

### -|-|-|-|-|-|-|-|-|-|-|- Answer 1a -|-|-|-|-|-|-|-|-|-|-|-

```{r}
require(DS705data)
data("anesthesia")
tbl <- table(anesthesia$anesthetic,anesthesia$nausea)
tbl
chisq.test(tbl)

```

Based on the $\chi$-squared test we can reject the null hypothesis that nausea is and anesthetics are not related.

### Part 1b

Obtain the output from R (including the Wald tests for coefficients - so use "summary" function) for the logistic regression model with nausea as the dependent variable and the type of anesthetic as the predictor variable.

### -|-|-|-|-|-|-|-|-|-|-|- Answer 1b -|-|-|-|-|-|-|-|-|-|-|-

```{r}
nausea.out <- glm(nausea~anesthetic,family="binomial", data = anesthesia)
summary(nausea.out)

```


### Part 1c  

What is the outcome of the hypothesis test that the coefficient of **anesthetic** is "zero" vs "not zero" at a 5% level of significance? (use the Wald test from the R output from the logistic regression you performed)

### -|-|-|-|-|-|-|-|-|-|-|- Answer 1c -|-|-|-|-|-|-|-|-|-|-|-

Based on a 5% level of significance the coefficient looks to be non-zero.

### Part 1d

Convert the estimated coefficient of **anesthesia** to an odds ratio and interpret it in the context of the problem.

### -|-|-|-|-|-|-|-|-|-|-|- Answer 1d -|-|-|-|-|-|-|-|-|-|-|-

```{r}
exp(-1.5261)

```

The odds of having nausea after surgery after taking anesthesia B are 21.7% as large as the odds of having nausea after surgery after taking anesthesia A, given all other variables are held constant.

### Part 1e

Install the package "mosaic" (if you don't have it installed already), then load it.  Use the oddsRatio function to compute the odds ratio for having nausea for anesthetic A vs B.  You may have to refer back to Week 8 for details on odds ratios and the oddsRatio function in R.

### -|-|-|-|-|-|-|-|-|-|-|- Answer 1e -|-|-|-|-|-|-|-|-|-|-|-

```{r}
#install.packages("mosaic")
require(mosaic)
oddsRatio(tbl)
```
4.6 to 1 for nausea if taking anesthesia A as compared to B.

### Part 1f

When logistic regression coefficients are negative, the interpretation sometimes has more impact when we switch the perspective and use the reciprocal of the exponentiated coefficient.  Find the odds ratio for having nausea for anesthetic B compared to anesthetic A (i.e. take the reciprocal of the odds ratio you computed in part **1d**). 

Interpret this odds ratio in the context of the problem.

### -|-|-|-|-|-|-|-|-|-|-|- Answer 1f -|-|-|-|-|-|-|-|-|-|-|-

```{r}
exp(1.5261)

```

Anestethic A is 4.6 more likely to cause nausea as compared to Anestethic B.

### Part 1g
    
Compute the predicted probability of a reconstructive knee surgery patient having nausea after surgery when anesthetic A was used.
    
### -|-|-|-|-|-|-|-|-|-|-|- Answer 1g -|-|-|-|-|-|-|-|-|-|-|-

```{r}
newdata <- data.frame(anesthetic='A')
predict(nausea.out,newdata,type="response")


```

----

### Part 1h
    
Compute a 95% confidence interval for the predicted probability of a reconstructive knee surgery patient having nausea after surgery when anesthetic A was used.
    
### -|-|-|-|-|-|-|-|-|-|-|- Answer 1h -|-|-|-|-|-|-|-|-|-|-|-

```{r}
out <- predict(nausea.out, newdata, se.fit=TRUE)
C = .95 # define the level of confidence
crit = qnorm(1-(1-C)/2) # get the appropriate critical value
lower = exp(out$fit-crit*out$se.fit)/(1+exp(out$fit-crit*out$se.fit))
upper = exp(out$fit+crit*out$se.fit)/(1+exp(out$fit+crit*out$se.fit))
c(lower,upper)


```

## Exercise 2

Continue using the anesthesia.rda data set to do the following.

### Part 2a

Obtain the output from R (including the Wald tests for coefficients - so use "summary" function) for the logistic regression model with nausea as the dependent variable and the amount of pain medication taken as the predictor variable.  

At $\alpha = 0.05$, is there a statistically significant relationship between nausea and the amount of pain medication taken?

### -|-|-|-|-|-|-|-|-|-|-|- Answer 2a -|-|-|-|-|-|-|-|-|-|-|-

```{r}
nausea.out2 <- glm(nausea~painmed,family="binomial", data = anesthesia)
summary(nausea.out2)

```

Based on a 5% level of significance the coefficient for painmed looks to be non-zero.

### Part 2b  

Convert the estimated coefficient of **painmed** to an odds ratio and interpret it in the context of the problem.

### -|-|-|-|-|-|-|-|-|-|-|- Answer 2b -|-|-|-|-|-|-|-|-|-|-|-

```{r}
exp(0.0375)

```

The odds of having nausea increases by 1.04 for each
additional unit of pain meds.

### Part 2c

Compute the predicted probabilities of a reconstructive knee surgery patient having nausea in the recovery time after surgery for when 50 units of pain medication are used and also for when 100 units of pain medication are used.

Comment on these two probabilities.

### -|-|-|-|-|-|-|-|-|-|-|- Answer 2c -|-|-|-|-|-|-|-|-|-|-|-

```{r}
newdata2 <- data.frame(painmed=c(50,100))
predict(nausea.out2,newdata2,type="response")

```

With 50 units of painmeds there is a 23.4% chance of the patient becoming nauseated, and with 100 units of painmeds there is a 66.6% chance of the patient becoming nauseated.
